---
title: 'Do More People Die of Extreme Heat or Excessive Cold?'
author: "Jim Callahan"
date: "September 25, 2015"
output:
  pdf_document: default
  html_document:
    keep_md: yes
---
#### Synopsis:
The question of **"Do More People Die of Extreme Heat or Excessive Cold?"** is 
addressed using the **"R"** statistical language and data from the 
**National Oceanic and Atmospheric Administration (NOAA) "Storm Data" database**.
To put the question in perspective we first look at what types of storms cause
the most deaths and property damage.

#### Data Processing  
The **R** language is an open source version of the the **S** language developed 
at Bell Labs during the "golden age" that also produced the **Unix** operating system
and the **C** programming language. See:  <https://www.R-project.org/>


Unlike **C**, **R** is an interpretive command language where the user types commands
at the command line and gets an immediate response:

```{r}
2+2

sqrt(25)

GaussDidThisInHisHeadInElemetarySchool <- sum(1:100)
print(GaussDidThisInHisHeadInElemetarySchool)
```

The last of these three examples is a problem solved by famous mathematician
**Carl Friedrich Gauss (1777-1855)** while he was an eight year old child math prodigy 
in elementary school. He solved it in his head, amazing his teacher. 
For more of the story see: "Clever Carl" <http://nrich.maths.org/2478/index?nomenu=1>`

For those of us who are not (child or adult) math prodigies we can solve the problem 
with **R** by typing **sum(1:100)** at the command line. The **"<-"** assigns
the result of the function to the variable name on the left.

The command line commands can be combined in simple text files **"scripts"** or combined
with compliled programs (compiled in **FORTRAN**, **C** or **C++**), data and 
documentation to form complete  **"packages"**.

Many open source statisical **"packages"** have been written in **R** making the 
complete system, the base language plus the optional downloadable statistical packages 
more powerful than proprietary statistical systems such as **SAS** or **SPSS** which, 
depending on the license, base price can be as much as $9,000 a seat.
See: <http://www.sas.com/store/products-solutions/cSoftware-p1.html>

When we load the **NOAA "Storm Data" data file** into the **R** statistical system,
we will also be using the **"<-"** to assign the result to the variable name on
the left.

```{r}
filename <- "~/GitHub/RepData_PeerAssessment2/data/repdata%2Fdata%2FStormData.csv.bz2"
NOAA <- read.csv(filename, 
                 stringsAsFactors = FALSE )
```


```{r}
# Select Columns of interest:
#        Primary key:    REFNUM
#        Date and Time:  BGN_DATE, BGN_TIME, TIME_ZONE,
#        Location:       STATE, COUNTY, COUNTY_END, LATITUDE, LONGITUDE,
#        Type of Storm:  EVTYPE,
#        Damage:         PROPDMG, PROPDMGEXP, CROPDMG, CROPDMGEXP,
#        Casualties:     FATALITIES, INJURIES

ColumnSubset <- c("REFNUM", "BGN_DATE", "BGN_TIME", "TIME_ZONE",
                  "STATE", "COUNTY", "COUNTY_END", "LATITUDE", "LONGITUDE", 
                  "EVTYPE", "PROPDMG", "PROPDMGEXP", "CROPDMG", "CROPDMGEXP", 
                  "FATALITIES", "INJURIES")

storms <- NOAA[, ColumnSubset]
str(storms)

#### Select Rows of interest 
#### Select years of interest (Since 2001 "21st Century Storms" or "last 25 years")

#### Step 1: Create a year variable from date
#### Read in the date.
#### NOAA's Time and time zone are in separate variables that we will ignore.
datetime = as.POSIXct(storms$BGN_DATE, "%m/%d/%Y %H:%M:%S", tz = "")
storms$year <- format(datetime, "%Y")

#### Step 2: Filter by year: "21st Century Storms"
storms <- storms[storms$year >= "2001", ]

#### Remove the original NOAA database
rm(NOAA)

```

```{r}
#### Scale Property Damage and Crop Damage by thouands and millions
# Have to create series before using with()
storms$PropertyDamage <- storms$PROPDMG   # Copy data; retain orginal intact

storms$PropertyDamage <- ifelse(storms$PROPDMGEXP == "K", 
                                storms$PropertyDamage * 1000, 
                                storms$PropertyDamage)

storms$PropertyDamage <- ifelse(storms$PROPDMGEXP == "M", 
                                storms$PropertyDamage * 1000*1000, 
                                storms$PropertyDamage)


storms$CropDamage     <- storms$CROPDMG   # Copy data; retain orginal intact

storms$CropDamage     <- ifelse(storms$CROPDMGEXP == "K", 
                                storms$CropDamage * 1000, 
                                storms$CropDamage)

storms$CropDamage     <- ifelse(storms$CROPDMGEXP == "M", 
                                storms$CropDamage * 1000*1000, 
                                storms$CropDamage) 
```


#### Tabulate Fatalities and Damage by Storm Type and Year
```{r}
#### Tabulate Fatalities and Damage by Storm Type and Year
#### Userful R commands include table(), xtabs(), ftable()  or aggregate()
#### This use of aggregate() is based on Jared Lander's "R for Everyone" page 123 
#### where he uses aggregate() on the diamonds data set from the ggplot2 package.
# Template: aggregate(formula, data, FUN, ..., subset, na.action = na.omit)
StormTot <- aggregate(
    formula = cbind(FATALITIES, INJURIES, PropertyDamage, CropDamage) ~ EVTYPE + year, 
    data    = storms, 
    FUN     = sum)

# Round off total to nearest dollar.
StormTot$PropertyDamage <- round(StormTot$PropertyDamage, digits = 0)
StormTot$CropDamage     <- round(StormTot$CropDamage, digits = 0)

#### Rename and put variables in logical order.
#### Order of columns:
StormTot <- StormTot[ , c("year", "EVTYPE", "FATALITIES", "INJURIES", 
                          "PropertyDamage", "CropDamage") ]

#### The variables in "stormtab" have been aggregated by type of storm and year
#### and thus the NOAA supplied names reflect the origin of the variable
#### but not its current content, so it is appropriate to rename the variables
#### for display.
ColumnNames <- c("Year", "StormType", "Fatalities", "Injuries", "PropertyDamage", "CropDamage")
colnames(StormTot) <- ColumnNames
```

#### Results
#### Health Impact
```{r}
# What type of storms caused the most fatalities in the most recent year (2011)?
MostRecentYear <- max(StormTot$Year)
StormsRankYear <- StormTot[StormTot$Year == MostRecentYear, ]

# This is the sort -- rank by fatalities in the stormrankyear
DeadlyStorms <- StormsRankYear[order(-StormsRankYear$Fatalities, na.last = NA), ]

# Renumber the rows
RowNames <- as.character(1:nrow(DeadlyStorms))
rownames(DeadlyStorms) <- RowNames

# head(DeadlyStorms, 25)
print( DeadlyStorms[1:25, 
         c("StormType", "Fatalities", "Injuries", "PropertyDamage", "CropDamage")] 
      )
```

#### Property Damage
```{r}
# What type of storms caused the most property damage in the most recent year (2011)?
DamageStorms <- StormsRankYear[order(-StormsRankYear$PropertyDamage, na.last = NA), ]

# Renumber the rows
RowNames <- as.character(1:nrow(DamageStorms))
rownames(DamageStorms) <- RowNames

# head(DamageStorms, 25)
print( DamageStorms[1:25, 
         c("StormType", "PropertyDamage", "CropDamage", "Fatalities", "Injuries")] 
      )
```


Event Descriptions indicating Hot or Cold Weather  
```
COLD  
COLD/WIND CHILL  
COLD WIND CHILL TEMPERATURES  
EXCESSIVE HEAT  
EXTREME COLD  
EXTREME COLD/WIND CHILL  
EXTREME WINDCHILL  
FREEZE  
FREEZING DRIZZLE  
FREEZING RAIN  
FROST/FREEZE  
HARD FREEZE  
HEAT  
RECORD WARMTH  
UNSEASONABLY COLD  
UNSEASONABLY COOL  
UNSEASONABLY WARM  
UNUSUALLY COLD  
WINTER STORM  
WINTER WEATHER  
WINTER WEATHER/MIX  
```

# END