---
title: 'Most Lethal and Deadly Storm Types in 2011'
author: "Jim Callahan"
date: "September 26, 2015"
output:
  pdf_document: default
  html_document:
    keep_md: yes
---
#### Synopsis:   

This paper is in response to **two questions:**  

> **1. Across the United States, which types of events (as indicated in the EVTYPE variable) 
> are most harmful with respect to population health?** According to the 2011 data: Tornadoes.
>
> **2. Across the United States, which types of events have the greatest economic consequences?**
>  According to the 2011 data: Floods.


These questions are addressed using the **"R"** statistical language and data from the 
**National Oceanic and Atmospheric Administration (NOAA) "Storm Data" database**.


#### Data Processing   

The **R** language is an open source version of the the **S** language developed 
at Bell Labs during the "golden age" that also produced the **Unix** operating system
and the **C** programming language. See:  <https://www.R-project.org/> for the **R**
language and for the history of S and R languages, see:
<https://www.r-project.org/nosvn/conferences/useR-2006/Slides/Chambers.pdf> ,  

<http://blog.revolutionanalytics.com/2014/01/john-chambers-recounts-the-history-of-s-and-r.html>   

and <https://www.stat.auckland.ac.nz/~ihaka/downloads/Massey.pdf>  


Unlike **C**, **R** is an interpretive command language where the user types commands
at the command line and gets an immediate response:

```{r}
2+2

sqrt(25)

GaussDidThisInHisHeadInElemetarySchool <- sum(1:100)
print(GaussDidThisInHisHeadInElemetarySchool)
```

The last of these three examples is a problem solved by famous mathematician
**Carl Friedrich Gauss (1777-1855)** while he was an eight year old child math prodigy 
in elementary school. He solved it in his head, amazing his teacher. 
For more of the story see: "Clever Carl" <http://nrich.maths.org/2478/index?nomenu=1>`

For those of us who are not (child or adult) math prodigies we can solve the problem 
with **R** by typing **sum(1:100)** at the command line. The **"<-"** assigns
the result of the function to the variable name on the left.

The command line commands can be combined in simple text files **"scripts"** or combined
with compliled programs (compiled in **FORTRAN**, **C** or **C++**), data and 
documentation to form complete  **"packages"**.

Many open source statisical **"packages"** have been written in **R** making the 
complete system, the base language plus the optional downloadable statistical packages 
competitive with traditional statistical systems such as **SAS** or **SPSS**.   

When we load the **NOAA "Storm Data" data file** into the **R** statistical system,
we will also be using the "**<-**" to assign the result to the variable name on
the left.

```{r}
filename <- "~/GitHub/RepData_PeerAssessment2/data/repdata%2Fdata%2FStormData.csv.bz2"
NOAA <- read.csv(filename, 
                 stringsAsFactors = FALSE )
```


```{r}
# Select Columns of interest:
#        Primary key:    REFNUM
#        Date and Time:  BGN_DATE, BGN_TIME, TIME_ZONE,
#        Location:       STATE, COUNTY, COUNTY_END, LATITUDE, LONGITUDE,
#        Type of Storm:  EVTYPE,
#        Damage:         PROPDMG, PROPDMGEXP, CROPDMG, CROPDMGEXP,
#        Casualties:     FATALITIES, INJURIES

ColumnSubset <- c("REFNUM", "BGN_DATE", 
                  "STATE", "COUNTY", "COUNTY_END", "LATITUDE", "LONGITUDE", 
                  "EVTYPE", "PROPDMG", "PROPDMGEXP", "CROPDMG", "CROPDMGEXP", 
                  "FATALITIES", "INJURIES")

storms <- NOAA[, ColumnSubset]
str(storms)

# Select Rows of interest 
# Select years of interest (Since 2001 "21st Century Storms" or "last 25 years")

# Step 1: Create a year variable from the date;
datetime = as.POSIXct(storms$BGN_DATE, "%m/%d/%Y %H:%M:%S", tz = "")
storms$year <- format(datetime, "%Y")

# Step 2: Filter by year: "21st Century Storms"
storms <- storms[storms$year >= "2001", ]

# Garbage Collection: Remove the original NOAA database from memory
rm(NOAA)

```

We can't simply sum the property damage variable, because some values are in thousands ("K")
and others are in millions ("M"). So, we need to rescale the variables using the appropriate
multipliers. The same goes for the crop damage variable.

```{r}
# Scale Property Damage and Crop Damage by thouands and millions
storms$PropertyDamage <- storms$PROPDMG   # Copy data; retain orginal intact

storms$PropertyDamage <- ifelse(storms$PROPDMGEXP == "K", 
                                storms$PropertyDamage * 1000, 
                                storms$PropertyDamage)

storms$PropertyDamage <- ifelse(storms$PROPDMGEXP == "M", 
                                storms$PropertyDamage * 1000*1000, 
                                storms$PropertyDamage)


storms$CropDamage     <- storms$CROPDMG   # Copy data; retain orginal intact

storms$CropDamage     <- ifelse(storms$CROPDMGEXP == "K", 
                                storms$CropDamage * 1000, 
                                storms$CropDamage)

storms$CropDamage     <- ifelse(storms$CROPDMGEXP == "M", 
                                storms$CropDamage * 1000*1000, 
                                storms$CropDamage) 
```


#### Fatalities and Damage by Storm Type and Year   

NOAA's "Storm Data" database is a comprehensive listing of major storm events.
Each tornado is a seperate event and tornados that cross state lines may count as
two tornado events. So, to compare the annual impact of "tornadoes" to the annual impact
of floods, we have to add up all of the data from the tornado events in a given year.
It is helpful to sum by year, because many managers are used to annual summaries. 
Moreover, while human lives are comparable, it is more problematic to add together damage 
estimates from the 1950s and 1960s when houses might cost in the tens of thousands of dollars
to damage estimates from the current century when the cost of houses are measured
in the hundreds of thousands of dollars. While it is simple to multiply by a Consumer Price Index (CPI) or a more specialized housing price index, given the magnitude of the price changes
the choice of index could by itself could distort the analysis.
```{r}
# Tabulate Fatalities and Damage by Storm Type and Year
# Userful R commands include table(), xtabs(), ftable()  or aggregate()
# This use of aggregate() is based on Jared Lander's "R for Everyone" page 123 
# where he uses aggregate() on the diamonds data set from the ggplot2 package.
# Template: aggregate(formula, data, FUN, ..., subset, na.action = na.omit)
StormTot <- aggregate(
    formula = cbind(FATALITIES, INJURIES, PropertyDamage, CropDamage) ~ EVTYPE + year, 
    data    = storms, 
    FUN     = sum)

# Round off total to nearest dollar
# because the estimates are not accurate to nearest penny.
StormTot$PropertyDamage <- round(StormTot$PropertyDamage, digits = 0)
StormTot$CropDamage     <- round(StormTot$CropDamage, digits = 0)

#### Rename and put variables in logical order.
#### Order of columns:
StormTot <- StormTot[ , c("year", "EVTYPE", "FATALITIES", "INJURIES", 
                          "PropertyDamage", "CropDamage") ]

#### The variables in "stormtot" have been aggregated by type of storm and year
#### and thus the NOAA supplied names reflect the origin of the variable
#### but not its current content, so it is appropriate to rename the variables
#### for display.
ColumnNames <- c("Year", "StormType", "Fatalities", "Injuries", "PropertyDamage", "CropDamage")
colnames(StormTot) <- ColumnNames
```

#### Results  

#### Health Impact  

This is in answer to the question:   

> **"Across the United States, which types of events (as indicated in the EVTYPE variable) 
> are most harmful with respect to population health?"**   

To prepare a ranking we sort the data and number the rows.
```{r}
# What type of storms caused the most fatalities in the most recent year (2011)?
MostRecentYear <- max(StormTot$Year)
StormsRankYear <- StormTot[StormTot$Year == MostRecentYear, ]

# This is the sort -- rank by fatalities in the stormrankyear
DeadlyStorms <- StormsRankYear[order(-StormsRankYear$Fatalities, na.last = NA), ]

# Renumber the rows
RowNames <- as.character(1:nrow(DeadlyStorms))
rownames(DeadlyStorms) <- RowNames
```

#### Health Impact in `r max(StormTot$Year)` alone  
```{r}
print( DeadlyStorms[1:25, 
         c("StormType", "Fatalities", "Injuries", "PropertyDamage", "CropDamage")] 
      )
```

#### Property Damage  

This is in answer to the question:   

> **Across the United States, which types of events have the greatest economic 
> consequences?**

Again, to prepare a ranking we sort the data and number the rows.
```{r}
# What type of storms caused the most property damage in the most recent year (2011)?
DamageStorms <- StormsRankYear[order(-StormsRankYear$PropertyDamage, na.last = NA), ]

# Renumber the rows
RowNames <- as.character(1:nrow(DamageStorms))
rownames(DamageStorms) <- RowNames
```

#### Property Damage  in `r max(StormTot$Year)` alone   
```{r}
print( DamageStorms[1:25, 
         c("StormType", "PropertyDamage", "CropDamage", "Fatalities", "Injuries")] 
      )
```


#### Do fatalities and storm damage always go together?   
```{r figure-plot, fig.width=4, fig.height=4}
# See Jared Lander's "R for Everyone: Advanced Analytics and Graphics", page 223
library(ggplot2)
ggplot(StormsRankYear, 
       aes(x=log(Fatalities), y=log(PropertyDamage))) + geom_point()

```

Based on the graph there seem to be many events with high property damage 
(the dots are clustered to the left of the graph, rather than along the diagonal 
or uniformly spread out) and low fatalities this may be due to evacuations 
which save human lives when property damage is unavoidable.